Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2019-10-09 doc/response.tex

\begin{equation*}
i_{t-\tau}(t) = \RR_0 i(t-\tau) g(\tau) S(t)
\end{equation*}

\revtext{p5-6: perhaps mention that $t, s$ is for calendar time and $\tau$ time relative to/from infection.}

We added the following sentence in the beginning of section 2 to clarify this:

``Hereafter, we use $t, s$ to represent calendar time and $\tau$ to
represent time since infection.''

@@

Fixed.

\revtext{p8, top and later: I have a problem with $N(a)$. Does your model allow for different number of acquaintances? But then your $\RR_0$ is no longer valid. Also, you are always assuming neighbours/acquaintances are chosen at random - please stress this more.}

We added the following sentences in the beginning of section 3 to make this clear:

``To explore the effect of multiple contacts on realized generation intervals, we relax our assumption that the population is homogeneous (and therefore do not rely on the classical definition of the basic reproductive number).
Instead, we assume that a disease spreads on a network;
infected individuals contact their ``acquaintances'' at random, but ``acquantanceships'' are predetermined by the network structure before the beginning of an epidemic.''

@@

\revtext{p8, bottom: I did not follow how you got $\hat g$. What is a, $N(a)$ and $f(a)$? Explain more.}

We tried to clarify this.

``... number of acquaintances $N(a)$ ...''

``The population-level egocentric kernel is found by integrating the individual-level egocentric over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.
Trapman \textit{et al.} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair ...''

@@

``When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$ through the Euler-Lotka equation.''

\revtext{In your analysis you assume that you observe the sampled contact traced generation times without error. It is known (e.g. your ref 18) that onset of symptoms rather than infection time are observed, and that often there are several potential infectors. You don't have to include this in your analysis, but at least it should be acknowledged.}

We added the following sentences:

``For simplicity, we assume that generation intervals are observed without error and assume that there is no under-reporting. 
We do not expect under-reporting to affect the inference of generation-interval distributions (unless there are systematic biases in the observation process).
On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number \cite{britton2019estimation}.
We do not pursue these direction in this study.''

@@

\revtext{p14, sec 7.2: the $S_i$ should vary over time. When is it measured in the algorithm?}

We added the following sentence:

``... (thus decreasing $S_i$ by 1 for all individuals $i$ that are acquaintances of the newly infected individual) ...''

### 2019-10-09 doc/network.tex
Autocommit (doc)
stuff

Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence}; 
making this distinction clear will be particularly useful when we discuss spatial components (see \fref{indpop}).
An individual-level intrinsic infection kernel $k(\tau; a)$ describes the rate at which an infected individual with ``aspect'' $a$ makes ``effective contacts'' (contacts which will cause infection if the contactee is susceptible).
Individual aspects may represent variation in the course of infection (e.g., duration of latent and infectious periods) and the level of infectiousness, which can depend both on biological infectiousness as well as contact patterns.
Hereafter, we use $t, s$ to represent calendar time and $\tau$ to
represent time relative to/from infection. \jd{Is this just time since infection?}

@@

Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \cite{trapman2016inferring}.
In this case, the infectiousness of an individual $\RR(a) = \int k(s; a) ds$ is the product of the number of acquaintances $N(a)$, which can vary among individuals, and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
for brevity, we do not pursue these directions here.
\jd{Confused now about $\delta$.}

\rev{I did not follow how you got $\hat g$. What is a, $N(a)$ and $f(a)$? Explain more.}
The population-level egocentric kernel is found by integrating the individual-level egocentric over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
\new{Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual been infected by the focal individual.}
Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair; for convenience, we will often omit ``population-level''.
Finally, the link between the growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
\begin{equation}
\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

@@

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap}
and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
\new{
For simplicity, we assume that \emph{all} generation intervals (no under-reporting) are observed without error.
In practice, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number \cite{britton2019estimation}.
We do not pursue these direction in this study.
} \jd{We should disentangle under-reporting from g-i estimation.}

>>>>

Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence}; 
making this distinction clear will be particularly useful when we discuss spatial components (see \fref{indpop}).
An individual-level intrinsic infection kernel $k(\tau; a)$ describes the rate at which an infected individual with ``aspect'' $a$ makes ``effective contacts'' (contacts which will cause infection if the contactee is susceptible).
Individual aspects may represent variation in the course of infection (e.g., duration of latent and infectious periods) and the level of infectiousness, which can depend both on biological infectiousness as well as contact patterns.
Hereafter, we use $t, s$ to represent calendar time and $\tau$ to
represent time since infection.

@@

Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \cite{trapman2016inferring}.
In this case, the infectiousness of an individual $\RR(a) = \int k(s; a) ds$ is the product of the number of acquaintances $N(a)$, which can vary among individuals, and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
for brevity, we do not pursue these directions here.

\rev{I did not follow how you got $\hat g$. What is a, $N(a)$ and $f(a)$? Explain more.}
The population-level egocentric kernel is found by integrating the individual-level egocentric over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) da,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
\new{Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.}
Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair; for convenience, we will often omit ``population-level''.
Finally, the link between the growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
\begin{equation}
\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

@@

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap}
and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
\new{
For simplicity, we assume that generation intervals are observed without error and assume that there is no under-reporting. 
We do not expect under-reporting to affect the inference of generation-interval distributions (unless there are systematic biases in the observation process).
On the other hand, it is difficult to measure generation intervals precisely because (1) infection events are often unobserved and (2) there may be multiple potential infectors; these factors can introduce biases to the estiamtes of the reproduction number \cite{britton2019estimation}.
We do not pursue these direction in this study.
}
